---
title: "setup"
author: "Givanna Putri"
date: "2025-10-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Please ensure you follow the instructions below **prior** to attending the workshop.

## Data download

Data for all the workshop exercises can be downloaded from OneDrive at this link: https://wehieduau-my.sharepoint.com/:f:/g/personal/putri_g_wehi_edu_au/IgDBDuhSSGbgTIDtq8Q1S9kIASL6kAVgwT0HbZ-JtQrAlwk?e=K50aHf.

Create a folder somewhere on your computer (e.g. 
`C:/2025_cytoconnect_spatial_workshop/` or `/Users/yourname/2025_cytoconnect_spatial_workshop/`
) to store all the files related for this workhop.
Create a `data` folder within it and download all the data files into that folder.

## Software installation

You need to download the following softwares and install them on your computer prior to the workshop:
- Qupath 0.6.0: https://qupath.readthedocs.io/en/stable/index.html
- ImageJ: https://imagej.net/ij/
- DeepCell ImageJ plugin: https://github.com/vanvalenlab/kiosk-imageJ-plugin
- R: https://www.r-project.org/
- RStudio: https://posit.co/download/rstudio-desktop/

After installing R and RStudio, please install the required R packages.
You can do this by running the code blocks below in RStudio.

Please install the packages that we will be using in the workshop by running
the code block below.

```{r, eval=FALSE}

cran_packages <- c(
  "Seurat", "ggplot2", "scales", "patchwork", "qs2",
  "viridis", "pak", "here", "tidyverse", "uwot", "pheatmap"
)

install.packages(cran_packages)

# RCTD only available on Github
pak::pkg_install("dmcable/spacexr")

bioc_packages <- c()
if(length(bioc_packages) > 0) {BiocManager::install(bioc_packages)}

```

Running this chunk will let you know if the packages have been installed properly. 

```{r eval=FALSE}
checkSetup <- function() {
  library(cli)
  cat("\n--------------------------------------\n")
  cat(style_bold(col_magenta("\n***Installing General Packages***\n\n")))
  not <- c(); not2 <- c()
  packages1 <- c(cran_packages, bioc_packages)#, "Test")
  for (i in 1:length(packages1)){
    if(requireNamespace(packages1[i], quietly = TRUE)==F) {
      cat(paste(style_bold(col_red(packages1[i])), "has not been installed\n"))
      not <- c(not,i)
    } else {
      suppressWarnings(suppressMessages(library(as.character(packages1[i]), character.only = TRUE)))
      cat(col_yellow(packages1[i]), "is loaded!\n")
    }
  }
  cat("\n--------------------------------------\n")
  if (length(not) > 0){
    cat(style_bold(bg_red("\n  **IMPORTANT**  ")),
        style_bold(col_yellow("\n\nYou need to install: \n")),
        paste(paste(c(packages1[not]), collapse=", ")),
        "\n\n--------------------------------------",
        "\n\n Use:\n - install.packages(),\n - BiocManager::install() or, \n - use Google to find installation instructions.\n\n", style_bold(col_green("Then run this function again!\n\n")))
  } else {
    cat("",col_green(style_bold("\n All packages are loaded!\n\n Happy Coding! :)\n\n")))
  }
}
checkSetup()
```

If you have any problems, please reach out to either:

- [Thomas O'Neil](mailto:thomas.oneil@sydney.edu.au)
- [Givanna Putri](mailto:putri.g@wehi.edu.au)

## Packages

Please install the packages that we will be using in the workshop by running
the code block below.

```{r, eval=FALSE}

cran_packages <- c(
  "BiocManager", "Seurat", "ggplot2", "scales", "patchwork", "qs2",
  "viridis", "pak", "here", "hdf5r", "Rfast2"
)

install.packages(cran_packages)

# RCTD only available on Github
pak::pkg_install("dmcable/spacexr")

bioc_packages <- c()
if(length(bioc_packages) > 0) {BiocManager::install(bioc_packages)}

```

Running this chunk will let you know if the packages have been installed properly. 

```{r eval=FALSE}
checkSetup <- function() {
  library(cli)
  cat("\n--------------------------------------\n")
  cat(style_bold(col_magenta("\n***Installing General Packages***\n\n")))
  not <- c(); not2 <- c()
  packages1 <- c(cran_packages, bioc_packages)#, "Test")
  for (i in 1:length(packages1)){
    if(requireNamespace(packages1[i], quietly = TRUE)==F) {
      cat(paste(style_bold(col_red(packages1[i])), "has not been installed\n"))
      not <- c(not,i)
    } else {
      suppressWarnings(suppressMessages(library(as.character(packages1[i]), character.only = TRUE)))
      cat(col_yellow(packages1[i]), "is loaded!\n")
    }
  }
  cat("\n--------------------------------------\n")
  if (length(not) > 0){
    cat(style_bold(bg_red("\n  **IMPORTANT**  ")),
        style_bold(col_yellow("\n\nYou need to install: \n")),
        paste(paste(c(packages1[not]), collapse=", ")),
        "\n\n--------------------------------------",
        "\n\n Use:\n - install.packages(),\n - BiocManager::install() or, \n - use Google to find installation instructions.\n\n", style_bold(col_green("Then run this function again!\n\n")))
  } else {
    cat("",col_green(style_bold("\n All packages are loaded!\n\n Happy Coding! :)\n\n")))
  }
}
checkSetup()
```


## Folder structure

The code block below will create a bunch of folders in the location where
this script is stored.

```{r, eval=FALSE}
# This sets the working directory for all subsequent code chunks to be run
# to the location of this file.

raw_data_dir <- file.path(here::here(), "data", "visium", "raw")
data_dir <- file.path(here::here(), "data", "visium", "data")

if(!dir.exists(raw_data_dir)){dir.create(raw_data_dir, recursive = TRUE)}
if(!dir.exists(data_dir)){dir.create(data_dir, recursive = TRUE)}
```

## Data Download{.tabset}

Please source the code block below to load a function to check the integrity of the downloaded files.

```{r eval=FALSE}
check_md5 <- function(files_expected_md5sum) {
    # files_expected_md5sum must be a named vector where the name is the path to the file
    # and the value is the expected md5sum
    library(cli)
    library(tools)
    
    error <- combine_ansi_styles("red", "bold")
    
    need_redownload <- FALSE
    for (filename in names(files_expected_md5sum)) {
        actual_md5sum <- md5sum(filename)
        expected_md5sum <- files_expected_md5sum[filename]
        if (actual_md5sum != expected_md5sum) {
            cat(error(paste(
                filename, "is corrupted. Actual md5sum:", actual_md5sum, "!= Expected md5sum:", expected_md5sum, "\n"
                )
            ))
            
            need_redownload <- TRUE
        }
    }
    if (need_redownload) {
        cat(error("Some files are corrupted. Please redownload"))
    }
}
```

***Note***: if you encounter a timeout error (after 60 seconds) while downloading the files,
you can increase the timeout to say 6,000 seconds to give R more time
to download the files by running `options(timeout = max(6000, getOption("timeout")))`

### Visium

The code block below will download the count matrix and spatial information
required to load the Visium data in.

```{r eval=FALSE}
download.file("https://cf.10xgenomics.com/samples/spatial-exp/3.0.0/Visium_V2_Human_Colon_Cancer_P2/Visium_V2_Human_Colon_Cancer_P2_filtered_feature_bc_matrix.h5", 
              destfile = file.path(raw_data_dir, "Visium_V2_Human_Colon_Cancer_P2_filtered_feature_bc_matrix.h5"), 
              method = "curl")


download.file("https://cf.10xgenomics.com/samples/spatial-exp/3.0.0/Visium_V2_Human_Colon_Cancer_P2/Visium_V2_Human_Colon_Cancer_P2_spatial.tar.gz", 
              destfile = file.path(raw_data_dir, "spatial.tar.gz"), 
              method = "curl")
```

The code below will check the integrity of the downloaded files.
If you get any red warning, please re-download the files that were found to be corrupted using the code block above.

```{r eval=FALSE}

# check md5sum of the downloaded files.
md5sum_str <- c(
    "0a9733cf0dfd3dfaa3b6c1c5c24bcbed",
    "a32edc825ab7089fca7848fc80cd5113"
)
names(md5sum_str) <- c(
    file.path(raw_data_dir, "Visium_V2_Human_Colon_Cancer_P2_filtered_feature_bc_matrix.h5"),
    file.path(raw_data_dir, "spatial.tar.gz")
)
check_md5(md5sum_str)

```

If there are no warning from the code block above, proceed to untar the `spatial.tar.gz` file 
using the code block below.

```{r eval=FALSE}
untar(file.path(raw_data_dir, "spatial.tar.gz"), exdir = raw_data_dir)
```


#### SC reference data

TODO. Need to upload the qs2 file somewhere they can download.

#### Optional loupe browser file

The following file is optional and is only required if you would like to try out
the demo we will be showing on how to use loupe browser to make some QC step simpler.

```{r eval=FALSE}
download.file(
    "https://cf.10xgenomics.com/samples/spatial-exp/3.0.0/Visium_V2_Human_Colon_Cancer_P2/Visium_V2_Human_Colon_Cancer_P2_cloupe.cloupe",
    destfile = file.path(raw_data_dir, "Visium_V2_Human_Colon_Cancer_P2_cloupe.cloupe"),
    method = "curl"
)
md5sum_str <- c("c5726e203c21a2fc3432831ea7c74252")
names(md5sum_str) <- c(
    file.path(raw_data_dir, "Visium_V2_Human_Colon_Cancer_P2_cloupe.cloupe")
)
check_md5(md5sum_str)
```

Install Loupe Browser at this link:
https://www.10xgenomics.com/support/software/loupe-browser/latest

## Session Info

```{r}
sessionInfo()
```
